### 2023-10-26 submission/response_to_reviewers.Rmd

---
output: pdf_document
geometry: margin=1.0in
header-includes:
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

**Reviewer #1 (Comments for the Author):**

**Within this manuscript Pat Schloss reports on the use of rarefaction to effectively control unequal observation efforts in 16S rRNA gene sequencing due to differences in sequencing depth. This paper is in response to a controversial idea within the microbiome community about whether rarefaction or other methods of controlling sequencing depth during diversity analysis are most appropriate. Overall, this paper shows convincing results that rarefaction is the most appropriate method for controlling unequal sampling depths at the present time. It was a pleasure to read and overall, I was happy with the manuscript. However, I do have a few comments that should be addressed by the author.**

**Major comments:**

**1. I found the result of increasing CoV in Bray-Curtis distances with increasing sequencing depth to be an interesting result given that increased sequencing depth should generally result in better confidence of what is within the underlying population. I'm wondering if the CoV results are difficult to understand due to the changing number of samples included at each sampling depth. Could this analysis be redone using only samples with sequencing depths that are above the highest rarefaction value chosen so that sample number does not vary across depth (this could be done in a subset of the samples/datasets presented). If changing sample number is not to blame are there any further speculations as to what is going on here?**

**2. It is clear based on the results shown that treatment groups that are highly associated with variance in sequencing depth will result in high false positive rates during diversity analysis when rarefaction is not used. However, I would argue that the simulation used does not in general reflect many natural datasets. It would be uncommon for most natural datasets to vary by treatment group by splitting sequencing depth via the median. I am wondering if this analysis could be done with varying levels of association between treatment group and sequencing depth. This will give the readers a better idea of how large this issue is in actual practice as the true association for most datasets may lay somewhere between figures 2 and 3. This would also help address how much of a concern this may be for previously published papers that do not use rarefaction during diversity analysis.**

**3. Based on the results of this paper it seems the key difference that is being demonstrated in these simulations as opposed to other works is the use of true rarefaction rather than subsampling. Further comments on this in the discussion would be useful to highlight why these results differ so significantly from other work such as the paper by Martino et al., in 2019 (PMID: 30801021) that introduced using robust CLR measures for microbiome data (and is quickly becoming one of the preferred choices of microbiome scientists). Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Matino et al's findings on robust CLR measures. Overall, a clarification on how these two publications come to differing conclusions about best practices when comparing beta diversity would be helpful to a large audience within the microbiome field.**

**4. I am wondering whether any of the differences in the results derived from studies such as Martino et al., 2019 (PMID: 30801021), can be attributed to sequence processing pipeline. While the author makes convincing arguments that removal of low abundance sequences may not always be appropriate, I am wondering how it would impact the results shown. Especially given the widespread use of this practice in the current literature as indicated by the author in the discussion. This would be especially important given the strong language around this practice in the discussion without primary data to back it up within this manuscript.**

**5. One line 164 the report indicates finding differences in richness between samples that should have the same OTU counts given the way the enrichment sampling was done (increasing sequencing counts not raw number of OTUS in the sample). It would make sense that metrics that consider abundances to change but not richness. Comments on this would be appreciated. Furthermore, this section jumps around between two different models of enrichment (abundance enrichment vs. raw count enrichment). A clearer separation of these two analyses would be appreciated to make parsing the text easier as in some cases it is unclear which enrichment strategy was used for the result that is being talked about.**

**Minor comments:**

**1. Well, the current consensus is that sequence depth is mostly attributed to noise (Gloor et al., 2017. PMID: 29187837). It would be nice to comment on some recent papers that indicate that through certain methods of library preparation sequencing depth may be correlated with sample biomass and how this reflects with the current thoughts in this manuscript (Cruz et al., 2021. PMID: 33717032, Munich et al., 2022. PMID: 36396943).**

**2. It would be nice if confidence intervals could be included in the correlation plots in Figure 1. Furthermore, if understood correctly Figures 2-3 y-axis should be the mean false positive rate over 100 dataset simulations? If so a measure of data spread would be appropriate for these values.**

**3. While I agree that the rarefaction value chosen is not arbitrary and is usually a balance between keeping samples and having the highest sequencing depth possible (as indicated in the discussion of this manuscript). It would be nice to give stronger guidelines around choosing an appropriate value to highlight more robust ways of choosing rarefaction depths. One thought that came across my mind would be the use of CoV calculations from repeated subsampling (rarefaction) to determine appropriate sequencing depths. Or perhaps just a simple calculation of good's coverage at various sequence depths (similar to rarefaction plots used now)?**

**4. Based on these results and various studies on microbiome biases, alpha diversity metrics such as richness do not indicate the true richness of a sample due to a variety of issues included sequencing depth. A comment on this in the discussion would be helpful, for those that are less seasoned in the microbiome field.**

**Reviewer #2 (Comments for the Author):**

**This study addresses a controversy that has been playing out in the literature regarding the processing of 16S rRNA gene amplicon data. Some researchers contend that rarefaction (the random subsampling with replacement to account for differences in sequencing depth among samples) is undesireable because it "wastes data". Schloss does a comprehensive analysis of how rarefaction performs compared to other ways of processing the samples (including using raw counts). He compares both alpha and beta diversity metrics using both a mock dataset and a collection of published datasets from a range of ecosystems. In nearly all cases, rarefaction**

**My only "major" critique relates to the core purpose of the article: to convince readers that rarefaction is a better way to process data than those proposed by other researchers. After reading the article multiple times, I still don't yet comprehend why the other researchers came to their conclusions. At some level, they must have convinced themselves that the use of raw read count (or other methods) out-perform rarefaction. Did they really not do the same comparisons that are done here? Without reading those carefully, I can't tell. Schloss does mention that they removed rare OTUs in their analysis. Is this the main reason why they did not detect the superior performance of rarefaction? I think just a brief explanation for Schloss' interpretation of why his conclusions are different from McMurdle/Holmes would be even more convincing for readers who don't want to dig too deeply into other papers. There is some discussion in the introduction, but I'd like to see it in the actual discussion section, after the full presentation of the new analysis.**

**Kudos. Love the running title. Tell us how you really feel!**

**Minor comments**
**Line 69 - This implies that McMurdle/Holmes used an incorrect interpretation of differences between rarefying and rarefaction. This is an opportunity to clarify! Do you agree with their interpretation (but want to promote the use of rarefaction specifically)?**

**Line 72 - Do you mean that the reanalysis did not remove 15% of the samples?**

**Line 72 - Change "method" to "methods"**

**Line 78 - Did they use the term "toy" in their work? This comes off as unnecessarily snarky. Would it be appropriate to use "simulated" instead?**

**Line 94 - Delete one of the "from"s**

**Line 113 - I would say that the Raw and SRS normalized are completely insensitive to seq depth when using inverse Simpson, based on Fig 1A? Not sure this matters in the end, but be transparent in the explanation of the results.**

**Line 167 - Were the 3% of OTUs selected at random, or did you remove the rarest 3%?**

**Line 194 - Change "increased" to "increase"**

**Line 210 - Drop your excessive sig figs .... Use "101" fold**

**Line 281 - I think this should be "my personal process for selecting the rarefaction threshold"?**

### 2023-11-16 submission/response_to_reviewers.Rmd

**Within this manuscript Pat Schloss reports on the use of rarefaction to effectively control unequal observation efforts in 16S rRNA gene sequencing due to differences in sequencing depth. This paper is in response to a controversial idea within the microbiome community about whether rarefaction or other methods of controlling sequencing depth during diversity analysis are most appropriate. Overall, this paper shows convincing results that rarefaction is the most appropriate method for controlling unequal sampling depths at the present time. It was a pleasure to read and overall, I was happy with the manuscript. However, I do have a few comments that should be addressed by the author.**

**Major comments:**

**1. I found the result of increasing CoV in Bray-Curtis distances with increasing sequencing depth to be an interesting result given that increased sequencing depth should generally result in better confidence of what is within the underlying population. I'm wondering if the CoV results are difficult to understand due to the changing number of samples included at each sampling depth. Could this analysis be redone using only samples with sequencing depths that are above the highest rarefaction value chosen so that sample number does not vary across depth (this could be done in a subset of the samples/datasets presented). If changing sample number is not to blame are there any further speculations as to what is going on here?**

> I have added a set of panels to Figure 6 indicating the standard deviation across the samples. This hopefully makes it clear that the standard deviation between samples is decreasing, but not as fast as the mean distance between samples. On the whole, the COV values are fairly small. I have added text to this effect at LXXX.

@@

**2. It is clear based on the results shown that treatment groups that are highly associated with variance in sequencing depth will result in high false positive rates during diversity analysis when rarefaction is not used. However, I would argue that the simulation used does not in general reflect many natural datasets. It would be uncommon for most natural datasets to vary by treatment group by splitting sequencing depth via the median. I am wondering if this analysis could be done with varying levels of association between treatment group and sequencing depth. This will give the readers a better idea of how large this issue is in actual practice as the true association for most datasets may lay somewhere between figures 2 and 3. This would also help address how much of a concern this may be for previously published papers that do not use rarefaction during diversity analysis.**

> XXXUse overlapping IQR instead of perfect confounding.
> Currently have code/get_intraquantile_design.R set at 10%/90%; needs to be run
> on great lakes

**3. Based on the results of this paper it seems the key difference that is being demonstrated in these simulations as opposed to other works is the use of true rarefaction rather than subsampling. Further comments on this in the discussion would be useful to highlight why these results differ so significantly from other work such as the paper by Martino et al., in 2019 (PMID: 30801021) that introduced using robust CLR measures for microbiome data (and is quickly becoming one of the preferred choices of microbiome scientists). Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Martino et al's findings on robust CLR measures. Overall, a clarification on how these two publications come to differing conclusions about best practices when comparing beta diversity would be helpful to a large audience within the microbiome field.**

> XXXDo this.

@@

**4. I am wondering whether any of the differences in the results derived from studies such as Martino et al., 2019 (PMID: 30801021), can be attributed to sequence processing pipeline. While the author makes convincing arguments that removal of low abundance sequences may not always be appropriate, I am wondering how it would impact the results shown. Especially given the widespread use of this practice in the current literature as indicated by the author in the discussion. This would be especially important given the strong language around this practice in the discussion without primary data to back it up within this manuscript.**

> XXXResponse to previous point will help to address this. But the data processing pipeline won't really factor in since all methods used the same data for a dataset. The methods don't know what type of processing was done upstream. Could have used purely synthetic distributions.

@@

**1. Well, the current consensus is that sequence depth is mostly attributed to noise (Gloor et al., 2017. PMID: 29187837). It would be nice to comment on some recent papers that indicate that through certain methods of library preparation sequencing depth may be correlated with sample biomass and how this reflects with the current thoughts in this manuscript (Cruz et al., 2021. PMID: 33717032, Munich et al., 2022. PMID: 36396943).**

> Look at suggested papers; cite Gloor

**2. It would be nice if confidence intervals could be included in the correlation plots in Figure 1. Furthermore, if understood correctly Figures 2-3 y-axis should be the mean false positive rate over 100 dataset simulations? If so a measure of data spread would be appropriate for these values.**

> The caption for Figure 1 indicates that the size of the confidence interval was smaller than the plotting symbol. As stated in the captions for Figures 2 and 3, "The false positive rate was the number of times a dataset yielded a significant result."

@@

**3. While I agree that the rarefaction value chosen is not arbitrary and is usually a balance between keeping samples and having the highest sequencing depth possible (as indicated in the discussion of this manuscript). It would be nice to give stronger guidelines around choosing an appropriate value to highlight more robust ways of choosing rarefaction depths. One thought that came across my mind would be the use of CoV calculations from repeated subsampling (rarefaction) to determine appropriate sequencing depths. Or perhaps just a simple calculation of good's coverage at various sequence depths (similar to rarefaction plots used now)?**

> I'm reluctant to add more than what I already have at L281-289. In that text, I mention using Good's coverage values as a check to make sure that the level of coverage is acceptable.

@@

**4. Based on these results and various studies on microbiome biases, alpha diversity metrics such as richness do not indicate the true richness of a sample due to a variety of issues included sequencing depth. A comment on this in the discussion would be helpful, for those that are less seasoned in the microbiome field.**

> I have added a sentence to this effect in the Discussion at **LXXX**

### 2023-11-16 submission/manuscript.Rmd
Comment on increasing BC COV values

***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample). To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6). The richness values increased with sequencing effort as rare OTUs would continue be detected. In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort. This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs. Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances. Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable. In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths. Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase. Regardless, the coefficients of variation were relatively small. This analysis indicates that there are benefits to increased sequencing depths.

@@

The up to 100-fold difference in sample sizes is an unfortunate byproduct of how sequencing libraries are constructed. Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequences that allow them to later assign sequences back to the samples that they came from. When the PCR products are pooled, efforts are taken to pool the fragments in equimolar ratios. Researchers use one of two approaches. First, they often will quantify the concentration of DNA from each PCR and then pool DNA in the desired amounts. Alternatively, they may use normalization plates where each well can hybridize a uniform amount of DNA that is then eluted and pooled. Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture. In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons since those sequences will be removed [@Morris2013]. Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.

>>>>

***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample). To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6). The richness values increased with sequencing effort as rare OTUs would continue be detected. In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort. This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs. Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances. Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable. In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths. Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase slightly for samples where the standard deviation did not decrease as quickly as the mean distance between samples. Regardless, the coefficients of variation were relatively small. This analysis indicates that there are benefits to increased sequencing depths.

@@

The up to 100-fold difference in sample sizes is an unfortunate byproduct of how sequencing libraries are constructed. Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequences that allow them to later assign sequences back to the samples that they came from. When the PCR products are pooled, efforts are taken to pool the fragments in equimolar ratios. Researchers use one of two approaches. First, they often will quantify the concentration of DNA from each PCR and then pool DNA in the desired amounts. Alternatively, they may use normalization plates where each well can hybridize a uniform amount of DNA that is then eluted and pooled. Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture. In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons since those sequences will be removed [@Morris2013]. Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample. Furthermore, it is important to keep in mind that because of residual sequencing errors, PCR artifacts, and incomplete sequencing, it is not possible to obtain absolute quantification of alpha or beta diversity values. Instead, comparisons of these values need to be made on a relative basis.
